---
title: "Bayesian Adaptive Methods for Clinical Trials - Example 1"
output:
  pdf_document: default
  html_notebook: default
  html_document:
    df_print: paged
date: "`r format(Sys.time(), '%d %B, %Y')`"
latex_engine: xelatex 
---



At the first interim we have enrolled $n_1$ patients. Each patient has a successful outcome $Y_i = 1$ or failure $Y_i = 0$. Assuming independence of patients and constant $p$ we get $X = \sum(Y_i)$. If we have observed $n_1$ patients then $X_1 \sim Bin(n_1, p)$. Using a conjugate $Beta(a, b)$ prior for $p$ we get posterior $Beta(X_1 + a, n_1 - X_1 + b)$ for $p$. If our prior is uniform $Beta(1, 1)$ this gives us $Beta(X_1 + 1, n_1 - X_1 + 1)$.

Say we have the option of enrolling an additional $n_2$ patients. The current posterior serves as a prior. We combine the prior with the likelihood $X_2 \sim Bin(n_2, p)$ which gives us posterior 
$p \sim Beta(X_1 + X_2 + 1, n_1 - X_1 + 1 + n_2 - X_2)$. 

The predictive point of view argues that we should sample from the interim posterior giving values $p_j^*$ from which we should simulate $X^*_{2j}$ from the $Bin(n_2, p_j^*)$ likelihood. Using these simulated values we can then make some decisions on the merits of keeping the trial going or quitting because it appears like a futile effort.

In practice we want to estimate quantiles from the posterior predictive distributions and will continue with the trial if the probability that the distribution of a pre-specified lower bound is above a given threshold is reasonably high, e.g. $Pr(p_{0.025} > 0.85) > 0.8$. Let's look at an example.
 

```{r}
suppressMessages(library("rethinking"))
suppressMessages(library("tidyverse"))

# Imagine a preliminary study had n1 patients of which x1 were (successful) events.
n1 <- 117
x1 <- 110

# Assuming a uniform prior, the posterior for p based on the preliminary study is:
post_1 <- rbeta(1e4, shape1 = x1 + 1, shape2 = n1 - x1 + 1)

# Lets say we want to achieve at least an 85% success
sum(post_1 > 0.85)/length(post_1)

rethinking::dens(post_1, main = "Posterior from Interim 1")
abline(v = 0.85, col = "red")

```

The preliminary study looks promising in that this single look at the data suggests that there is a (nearly) 100% probability that our success rate is $\ge 0.85$, should we enrol more participants? Let's simulate this next stage of the trial and see if keeping it going makes sense. Let's say we could enrol another $n_2 = 200$ patients. To simulate we generate a 1000 separate trial populations each comprising 200 patients and each trial population will have the number of successes parameterised by the first interim posterior distribution (`post_1`).

```{r}
n2 <- 200

# Sample from post_1 to obtain the distribution of successful events.
n_trial_pops <- 1000
n_sims <- 1e4
x2 <- rbinom(n_trial_pops, n2, prob = post_1)

# This is the distribution of events across all of the simulated populations.
rethinking::simplehist(x2)
```

So typically, based on the histogram, about 185 to 190 of the 200 participants have a successesful outcome. We iterate through each of the simulated trial populations and generate the posterior distribution based on the number of successful events at the end of each study. Again, all we need to do is update the conjugate beta posterior. 

```{r}
post_2 <- numeric(n_sims * n_trial_pops)
dim(post_2) <- c(n_sims, n_trial_pops)

for(i in 1:length(x2)){
  post_2[,i] <- rbeta(n_sims, 
                shape1 = x1 + x2[i] + 1, 
                shape2 = n1 + n2 - x1 - x2[i] + 1)
}

```

Let's examine the posteriors arising from (a subset of) our simualted populations to get a sense of the heterogenity.

```{r}
df_post_2 <- as.data.frame(post_2) 

idx <- sample(1:1000, size = 10, replace = F)
df_post_2_sub <- df_post_2[,idx]
df_post_2_sub <- tidyr::gather(df_post_2_sub, "sim", "p")

ggplot(df_post_2_sub, aes(x = p, group = sim)) +
  geom_density()

```

Each trial produces an empirical distribution for $p^*_j$ and there is quite a bit of variation as seen above. We could then extract a lower bound (say the 2.5 percentile) from each of these posteriors.

```{r}
post_2_q_025 <- apply(post_2, 2, quantile, 0.025)
rethinking::dens(post_2_q_025)
```

Finally, we assess the probability that this quantile (that represents a lower bound arising from each of the simulated studies) will occur against our decision criteria - let's say we will not go ahead unless there is an 80% chance that our lower bound is going be greater than 0.85 success rate. For this particular scenario it looks like there is a good chance that the trial will be worth continuing so we enrol a further $n_2$ patients.

```{r}
# Descision threshold. 
# We compute Pr(p_0.025 > 0.85) = number of p^*_0.025 > 0.85 / N_rep

prob_p_025 <- sum(post_2_q_025 > 0.85) / length(post_2_q_025)

sprintf("Probability that 2.5%% quantile > 0.85 = %.3f ", prob_p_025)

ifelse(prob_p_025 > 0.8, 
       paste("Continue trial"), 
       paste("Discontinue trial")) 
```







